---
title: "Analysis of State Prison Inmates"
author: "Hannah V., Sophie K., Emily W., & Starr M."
date: "April 23, 2026"
logoleft_name: "bars.png"
logoright_name: "MH.png"
primary_colour: "#F87E1B"
title_textcol: "#FFDDBF"
body_textcol: "#913526"
institute: "University of Virginia"
body_bgcol: "#C4ECFF"
poster_height: "36in"
poster_width: "48in"
column_numbers: 4
font_family: Garamond
output:
  posterdown::posterdown_html
encoding: UTF-8
---
```{r setup, include = F}
options(encoding = "UTF-8")
library(tidyverse)
library(dplyr)
library(knitr)
library(caret)
library(kableExtra)
library(ggplot2)
library(htmltools)
library(broom)

```

# Introduction and Data 

**Background**<br>
Rule violations during incarceration reflect both inmate wellbeing and institutional safety. This project uses rule violations as an early signal of risk, shifting attention away from only post‑release outcomes toward what happens during prison sentences. Our goal is to show where mental health services and other supports could most reduce infractions and improve conditions for incarcerated people.

**Research Questions**<br>
We focus on two main questions:

<u>Which demographic, criminal history, and clinical factors are associated with higher infraction risk?</u>

<u>How do mental health diagnoses and in‑prison mental health treatment relate to that risk, and do current services appear to reach those at highest risk?</u>

**Data**<br>
We use the _2016 Survey of Prison Inmates (SPI), United States (ICPSR 37692)_, a public‑use survey of state prisoners. About 20,000 inmates were selected through a two‑stage probability sample to represent roughly 1.5 million state prisoners nationwide. Data come from in‑person, computer‑assisted interviews and linked records, with anonymity protections and a relatively high response rate.

<u>**Key Variables**</u><br>
Our analysis focuses on:

**Mental health**: number of mental health diagnoses (severity proxy) and indicators of receiving mental health treatment during incarceration.

**Substance use**: indicators of alcohol abuse and participation in alcohol treatment programs.

**Social and criminal history**: education, prior arrests, and other demographic and offense characteristics used as controls.


# Methods of Analysis {.mybreak}

BLAH BLAH BLAH


# Results {.mybreak}

BLAH BLAH BLAH

# Data visualizations 

```{r, echo = F, warning = F, message=F}

cleaned2<-read.csv('cleaned2.csv')


model3 <- glm(at_risk ~ `Years.of.Education` + `ADD.ADHD` + 
                Gender + offense_type + Children + `Times.Arrested` + 
                `Mental.Health.Treatment` + Alcoholism,
              family = binomial(link = "logit"), 
              data = cleaned2,
              na.action = na.exclude) 



cleaned2$binary_risk<-cleaned2$pred_risk_level
cleaned2$binary_risk[cleaned2$binary_risk=="High Predicted Risk"] <- 1
cleaned2$binary_risk[cleaned2$binary_risk=="Low Predicted Risk"] <- 0







cm_data <- data.frame(
  Classification = c("Predicted: Low Risk", "Predicted: High Risk"),
  `Observed: No Infraction (0)` = c("5,361", "1,755"),
  `Observed: At Risk (1)` = c("2,365", "5,969"),
  check.names = FALSE
)
cm <- confusionMatrix(as.factor(cleaned2$binary_risk), as.factor(cleaned2$at_risk))

kable(
  cm_data,
  booktabs = TRUE,
  caption = "Stage 1 Confusion Matrix",
  align = "lcc",
  escape = FALSE
) %>%
  footnote(
    general = "Overall Accuracy: 73.3% | Sensitivity: 75.3% | Specificity: 71.6% | P-Value: < .001 | No Information Rate (NIR): 53.9%",
    threeparttable = TRUE
  )
```


```{r, echo = F, warning = F, message=F, fig.width=12, fig.height=8, out.width="100%"}
temp_data <- data.frame(
  Times.Arrested = seq(0, 100, length.out = 100),
  Years.of.Education = mean(cleaned2$Years.of.Education, na.rm = TRUE),
  ADD.ADHD = 0,                   
  Gender = "Male",
  offense_type = "Property",
  Children = 1,
  Mental.Health.Treatment = 0,     
  Alcoholism = 1                  
)

temp_data$prob <- predict(model3, newdata = temp_data, type = "response")

ggplot(temp_data, aes(x = Times.Arrested, y = prob)) +
  geom_line(color = "#0077B5", size = 1.5) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Probability of being 'At Risk' due to Arrest History",
       subtitle = "Holding all other factors constant at the baseline (No ADHD, No Treatment,)",
       x = "Number of Times Arrested",
       y = "Predicted Probability") + 
  theme(text = element_text(size = 22))
  
```


```{r, echo = F, warning = F, message=F,  out.width="150%" }
stage1_table <- data.frame(
  Predictor = c(
    "Mental Health Treatment",
    "ADD/ADHD",
    "Times Arrested",
    "Children"
  ),
  "Odds Ratio" = c("9.0x", "3.8x", "1.01x", "0.91x"),
  Interpretation = c(
    "Highest impact group",
    "High clinical risk",
    "Risk increases per arrest",
    "Protective factor (9% lower risk)"
  ),
  check.names = FALSE
)

table_output <- kable(stage1_table,
      booktabs = TRUE,
      align = "lll",
      caption = "Stage 1: Key Predictors of At-Risk Classification") %>%
  kable_styling(
    full_width = TRUE,
    font_size = 30,
    bootstrap_options = c("striped", "hover")
  ) %>%
  row_spec(0, bold = TRUE, background = "#0077B5", color = "white") %>%
  row_spec(1, background = "#fff3cd") %>%  
  row_spec(2, background = "#39ae85") %>%
  row_spec(3, background = "#fff3cd") %>%
  row_spec(4, background = "#39ae85") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, color = "#0077B5", bold = TRUE)

div(style = "transform: scale(1.25); transform-origin: top left; width: 75%;",
    HTML(as.character(table_output)))


```

<br>

```{r, warning=F, message=F, echo=F,fig.width=11.5, fig.height=8 }
library(ggplot2)
library(dplyr)

coef_data <- data.frame(
  term = c( "Race: Black", "Race: Hispanic", "Race: Multiracial", "Race: Other", "Race: White", "Num. Diagnoses", "Armed Forces", "Times Arrested", "Years of Education", "ADD/ADHD", "Homeless", "Children", "Mental Health Treatment", "Alcoholism", "Alcohol/Drug Treatment", "Gender: Male", "Offense: Other", "Offense: Property", "Offense: Public Order", "Offense: Unknown", "Offense: Violent", "MH Treatment x Diagnoses", "Alcoholism x Drug Treatment"
  ),
  estimate = c(
    1.035531, 1.054957, 0.730125, 1.330617, 0.439531, 0.072018, -0.145850, 0.007912, -0.037718, 0.604736, -0.114640, -0.395354, 0.324830, 0.294180, 0.023718, -0.045294, 0.400541, -0.014625, -0.240554, -0.512680, 0.125330, -0.014916, -0.281589
  ),
  std.error = c(0.490403, 0.511488, 0.489379, 0.536754, 0.489300,0.026254, 0.119248, 0.001149, 0.011595,0.063758, 0.113859, 0.058881, 0.093639,0.081531, 0.074325, 0.062580,0.417414, 0.096488, 0.111770,0.133487, 0.085486,0.033334, 0.109103
  ),
  p.value = c( 0.034721, 0.039158, 0.135714, 0.013175, 0.369034, 0.006086, 0.221300, 5.76e-12, 0.001142, 2e-16, 0.314003, 1.89e-11, 0.000522, 0.000308, 0.749638, 0.469206, 0.337268, 0.879527, 0.031380, 0.000123, 0.142624, 0.654547, 0.009853
  )
) %>%
  mutate(
    conf.low = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    significant = p.value < 0.05,
    direction = case_when(
      estimate > 0 & significant ~ "Increases Risk",
      estimate < 0 & significant ~ "Decreases Risk",
      TRUE ~ "Non-Significant"
    ),
    highlight = term == "Children",
    term = reorder(term, estimate)
  )

ggplot(coef_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.7) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high, color = direction),
    height = 0.3, linewidth = 0.7
  ) +
  geom_point(
    aes(color = direction, size = highlight, shape = highlight)
  ) +
  geom_text(
    data = filter(coef_data, highlight),
    aes(label = "  ← Protective Factor"),
    hjust = 0, color = "#39ae85", fontface = "bold", size = 4.5
  ) +
  scale_color_manual(
    values = c(
      "Increases Risk"    = "#C0392B",
      "Decreases Risk"    = "#39ae85",
      "Non-Significant"   = "gray60"
    )
  ) +
  scale_size_manual(values = c("FALSE" = 3, "TRUE" = 5), guide = "none") +
  scale_shape_manual(values = c("FALSE" = 16, "TRUE" = 18), guide = "none") +
  labs(
    title    = "Landscape of Risk",
    subtitle = "Negative Binomial Model Coefficients with 95% Confidence Intervals",
    x = "Coefficient Estimate",
    y = NULL,
    color = NULL,
    caption = "Points in red increase infraction risk"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(linetype = "dotted", color = "gray85"),
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    axis.text.y = element_text(size = 11)
  )
      
```


# Interpretations {.mybreak}

BLAH BLAH BLAH
