---
title: "Technical Report"
author: "Emily Wu and Starr Mark"
date: "2026-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyr)
library(dplyr)
library(caret)
library(kableExtra)
library(MASS)
library(performance)
library(see)
library(ggplot2)

```


# Results

Our Final First Stage Model is as follows:

$P(At Risk) = \beta_0+\beta_1Years of Education+\beta_2ADD/ADHD+\beta_3Gender:Male+\beta_4Offense:Other+\beta_5Offense:Property+\beta_6Offense:Public Order+\beta_7Offense:Unknown+\beta_8Offense:Violent+\beta_9Children+\beta_{10}Times Arrested+\beta_{11}Mental Health Treatment+\beta_{12}MentalHealthTreatment$

#### FIXXXXXX LATER 

After conducting a Likelihood-Ratio Test on our model, we find that our final model has a deviance score of 4685 on 12 degrees of freedom and a p-value less than 0.001. This model was fit on $n=15,450$ observations with 1708 dropped/deleted observations. Over 90% are due to alcoholism.

This indicates our full model significantly outperformed the null model and the combination of clinical and demographic factors provides substantial predictive power for inmate infraction risk. 
Overall, our model had a 73% classification accuracy (classification of inmates who had infractions or not)
```{r, echo=F}
cleaned2<-read.csv('cleaned2.csv')


model3 <- glm(at_risk ~ `Years.of.Education` + `ADD.ADHD` + 
                Gender + offense_type + Children + `Times.Arrested` + 
                `Mental.Health.Treatment` + Alcoholism,
              family = binomial(link = "logit"), 
              data = cleaned2,
              na.action = na.exclude) 



cleaned2$binary_risk<-cleaned2$pred_risk_level
cleaned2$binary_risk[cleaned2$binary_risk=="High Predicted Risk"] <- 1
cleaned2$binary_risk[cleaned2$binary_risk=="Low Predicted Risk"] <- 0







cm_data <- data.frame(
  Classification = c("Predicted: Low Risk (0)", "Predicted: High Risk (1)"),
  `Observed: No Infraction (0)` = c("5,361", "1,755"),
  `Observed: At Risk (1)` = c("2,365", "5,969"),
  check.names = FALSE
)
cm <- confusionMatrix(as.factor(cleaned2$binary_risk), as.factor(cleaned2$at_risk))
conf_intervals <- suppressMessages(exp(confint(model3)))
vars <- c("ADD.ADHD", "GenderMale", "Mental.Health.Treatment")
conf_intervals <- suppressMessages(exp(confint(model3)))
odds_ratios <- suppressMessages(exp(coef(model3)))
vardf <- data.frame(
  Variable = c("ADD/ADHD", "Male Gender", "Mental Health Treatment"),
  Estimate = model3$coefficients[vars],
  OR = odds_ratios[vars],
  "Lower_CI" = conf_intervals[vars, 1],
  "Upper_CI" = conf_intervals[vars, 2],
  "Z_Value" = summary(model3)$coefficients[vars, "z value"],
  row.names = NULL
)
vardf$"95% Confint OR" <- paste0("(", round(vardf$Lower_CI, 2), " - ", round(vardf$Upper_CI, 2), ")")

acc  <- round(cm$overall["Accuracy"] * 100, 1)
nir  <- round(cm$overall["AccuracyNull"] * 100, 1)
pval <- signif(cm$overall["AccuracyPValue"], 3)




kable(
  cm_data,
  booktabs = TRUE,
  caption = "Stage 1 Confusion Matrix",
  align = "lcc",
  escape = FALSE
) %>%
  footnote(
    general = "Overall Accuracy: 73.3% | Sensitivity: 75.3% | Specificity: 71.6% | P-Value: < .001 | No Information Rate (NIR): 53.9%",
    threeparttable = TRUE
  )




```


The model diagnostics state that the most significant predictors are ADD/ADHD, Male Gender, Times Arrested, and Mental Health Treatment. Out of these, the predictors with the largest effect size are ADD/ADHD with 1.33, Male Gender with -0.938, and Mental Health Treatment with 2.201. 

```{r, echo = F}
kable(vardf[c(1, 2, 3, 6, 7)]
, caption = "Significant Predictors",
          threeparttable = TRUE
)
```


\newpage

The final Second Stage Model is as follows:

$ln(Infractions) = \beta_0+\beta_1Race:Black+\beta_2Race:Hispanic+\beta_3Race:Multiracial+\beta_4Race:Other+\beta_5Race:White+\beta_6NumberofDiagnoses+\beta_7ArmedForces+\beta_8TimesArrested+\beta_9YearsofEducation+\beta_{10}ADD/ADHD+\beta_{11}Homeless+\beta_{12}Children+\beta_{12}MentalHealthTreatment+\beta_{13}Alcoholism+\beta_{14}AlcoholTreatment+\beta_{15}Gender:Male+\beta_{16}+ \beta_{17}Offense:Other+\beta_{18}Offense:Property+\beta_{19}Offense:PublicOrder+\beta_{20}Offense:Unknown+\beta_{21}Offense:Violent+\beta_{22}NumberofDiagnoses*MentalHealthTreatment+\beta_{23}Alcoholism*Alcohol Treatment$

Comparing the 2 models using their goodness-of-fit statistics, it is seen that a negative binomial model is more favorable with a lower AIC and Overdispersion statistic $(\phi)$. The negative binomial model was fitted on the subpopulation of $n = 7,619$ of those rated as "High Predicted Risk." This was determined by our previous logistic model of those who had a higher probability of having an infraction than the median value of the data set, which was 0.497. 105 observations were not counted, over 90% were deleted due to Race being unavailable. 


```{r, echo=F}
high_risk_group <- cleaned2 %>% filter(pred_risk_level == "High Predicted Risk")
high_risk_group <- high_risk_group %>%
  mutate(Number.Infractions = ifelse(Number.Infractions < 0, NA, Number.Infractions))

stage2_model <- glm.nb(Number.Infractions ~ Race + num_diagnoses + Armed.Forces + +
                         Times.Arrested+ Years.of.Education+ ADD.ADHD + Homeless+ Children+ 
                         Mental.Health.Treatment+ Mental.Health.Treatment:num_diagnoses+ Alcoholism
                       + Alcohol.Drug.Treatment + Alcoholism:Alcohol.Drug.Treatment + Gender + offense_type, data = high_risk_group)


stage2_pois <- glm(Number.Infractions ~ Race + num_diagnoses + Armed.Forces + +
                         Times.Arrested+ Years.of.Education+ ADD.ADHD + Homeless+ Children+ 
                         Mental.Health.Treatment+ Mental.Health.Treatment:num_diagnoses+ Alcoholism
                       + Alcohol.Drug.Treatment + Alcoholism:Alcohol.Drug.Treatment + Gender + offense_type, family = "poisson", data = high_risk_group)


moddf<-data.frame(
  Model = c("Poisson","Negative Binomial"),
  AIC = c(stage2_pois$aic, stage2_model$aic),
  Deviance = c(stage2_pois$null.deviance-stage2_pois$deviance, stage2_model$null.deviance-stage2_model$deviance),
  df = c(23, 23),
  Overdispersion = c(sum(residuals(stage2_pois, type = "pearson")^2) / df.residual(stage2_pois), stage2_model$theta)
)

kable(moddf, caption = "Model Comparison",
          threeparttable = TRUE
)

```




In our model, the most significant predictors are ADD/ADHD, Times Arrested, Children, Mental Health Treatment, Alcoholism, and an Unknown Offense Type. The variables with the largest effect size were ADD/ADHD with a 0.605 Coefficient and Unknown Offense type with -0.513 coefficient. 


```{r, echo = F}
vardf2<- data.frame(
  Variable = c("ADD/ADHD", "Times Arrested", "Children", "Mental Health Treatment", "Alcoholism", "Offense Type: Unknown"),
  Estimate = c(stage2_model$coefficients["ADD.ADHD"],stage2_model$coefficients["Times.Arrested"],stage2_model$coefficients["Children"],stage2_model$coefficients["Mental.Health.Treatment"],stage2_model$coefficients["Alcoholism"],stage2_model$coefficients["offense_typeUnknown"] ),
  "Z-Value" = c(summary(stage2_model)$coefficients[11,3],summary(stage2_model)$coefficients[9,3],summary(stage2_model)$coefficients[13,3], summary(stage2_model)$coefficients[14,3], summary(stage2_model)$coefficients[15,3], summary(stage2_model)$coefficients[21,3]),

  "P-Value" = c(summary(stage2_model)$coefficients[11,4],summary(stage2_model)$coefficients[9,4],summary(stage2_model)$coefficients[13,4], summary(stage2_model)$coefficients[14,4], summary(stage2_model)$coefficients[15,4], summary(stage2_model)$coefficients[21,4]),
  row.names = NULL
)

kable(vardf2, caption = "Model Statistics",
          threeparttable = TRUE
)
```





```{r}
fittedN<-model.frame(stage2_model)


count_comp <- data.frame(
  Observed = fittedN$Number.Infractions,
  Fitted = predict(stage2_model)
) %>%
  # Round fitted values to nearest integer to compare counts
  mutate(Fitted = round(Fitted)) %>%
  gather(Type, Value) %>%
  filter(Value <= 10) # Filter to low counts (where most action happens)

# 2. Plot the distributions
ggplot(count_comp, aes(x = Value, fill = Type)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Observed vs. Fitted Counts (Negative Binomial)",
    x = "Number of Events",
    y = "Frequency"
  ) +
  theme_minimal()
check_predictions(stage2_model)
```







